---
title: "data_setup"
author: "Micah GVI"
date: "04/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r package load, warning=FALSE, message=FALSE}
library(BiocManager)
library(annotate)
library(ape)
library(ggtree)
library(seqinr)
library(Biostrings)
library(wesanderson)
library(viridis)
library(ggplot2)
library(stringi)
library(treeio)
library(reshape2)
library(phytools)
library(mapdata)
```

### Amino acid sequence code ####
```{r data import, echo=FALSE}
seq <- readAAStringSet("./input/master_aa_seqs.fa") #upload data file
seqDF <- data.frame(seq) #create a data frame with the data 
names(seqDF) <- c("seq") 
```


```{r}
#Extract sequence titles to format/add information
title = rownames(seqDF)
title = as.data.frame(title)

write.csv(names, "./input/title.csv", row.names=FALSE)

#Missing Vector and Country information is added in here using accession literature 
#File is re-saved as titlefinal

WNV_Variant = read.csv("./input/titlefinal.csv")

seqDF = cbind("WNV Variant" = WNV_Variant, seqDF)
rownames(seqDF) = seqDF[,1]
```


```{r}
## Data cleanup/removal of abberant seqs
# Check sequence lengths.
SeqLen<-as.numeric(lapply(seqDF$seq,nchar))
qplot(SeqLen, binwidth = 50)+theme_bw() # There are nine sequences that are drastically different lengths than the rest. I will remove these:
KeepSeq<-SeqLen>3000 #removing anything greater than 3000 in length 
seqDF<-seqDF[KeepSeq,] #re-writing the data frame with new wanted data 
```

### Isolating structural proteins.
```{r}
cpro_seq<- unlist(substr(seqDF$seq, start = 6, stop = 123))
cproAA<-sapply(cpro_seq,strsplit,split="")
names(cproAA) <- seqDF$Title
cproAA <- as.AAbin(cproAA)


prMpro_seq <- unlist(substr(seqDF$seq, start = 217, stop = 290))
prMproAA<-sapply(prMpro_seq,strsplit,split="")
names(prMproAA) <- seqDF$Title
prMproAA <- as.AAbin(prMproAA)


epro_seq <- unlist(substr(seqDF$seq, start = 598, stop = 791))
eproAA<-sapply(epro_seq,strsplit,split="")
names(eproAA) <- seqDF$Title
eproAA <- as.AAbin(eproAA)
```

  
### Function for Alignment and Distance Matrices
```{r DM function} 
DMfunction <- function(protein) {
  align<-muscle(protein,quiet=F)
  
  alignmentvisual <- image.AAbin(align, bg = (wes_palette("Darjeeling1")[5]), col = wes_palette("Darjeeling1"), ylab = "WNV Variants", xlab = "Sequence Length", show.labels = T, cex.lab = 0.2, legend = T, aa.cex = 0.8)
    DM<-dist.aa(align) # generate distance matrix based on AA sequences.
  DMmat<-as.matrix(DM) # convert to matrix format.

# Melt to a linear matrix and generate distance matrix visual.
DMlmat<-melt(DMmat) # Linear matrix.
ggplot(data=DMlmat,aes(x=Var1,y=Var2,fill=value)) + geom_tile() + scale_fill_gradientn(colours=c("white","blue","green","red")) + theme(text = element_text(size=5),
        axis.text.x = element_text(angle=90, hjust=1)) + ggtitle("C Protein Distance Matrix")
}
```

```{r}
DMfunction(cproAA)
DMfunction(prMproAA)
DMfunction(eproAA)
```

### Function for Alignment
I don't know why this function does not create the new DM objects to be used in code later... 
```{r DM function} 
AlignmentFunction <- function(protein) {
  align<-muscle(protein,quiet=F)
  alignmentvisual <- image.AAbin(align, bg = (wes_palette("Darjeeling1")[5]), col = wes_palette("Darjeeling1"), ylab = "WNV Variants", xlab = "Sequence Length", show.labels = T, cex.lab = 0.2, legend = T, aa.cex = 0.8)
  
#Generate distance matrix based on AA sequences
   if (protein == "cproAA"){
    CproDM<-dist.aa(align)
  }
  if (protein == "prMproAA"){
    prMproDM<-dist.aa(align)
  }
   if (protein == "eproAA"){
    eproDM<-dist.aa(align)
   }
}



```

```{r}
AlignmentFunction(cproAA)
AlignmentFunction(prMproAA)
AlignmentFunction(eproAA)
```

###Function for Distance Matrix 
```{r}
MatrixFunction <- function(DM) {
  DMmat<-as.matrix(DM) # convert to matrix format
  
# Melt to a linear matrix and generate distance matrix visual.
DMlmat<-melt(DMmat) # Linear matrix.
ggplot(data=DMlmat,aes(x=Var1,y=Var2,fill=value)) + geom_tile() + scale_fill_gradientn(colours=c("white","blue","green","red")) + theme(text = element_text(size=5),
        axis.text.x = element_text(angle=90, hjust=1)) + ggtitle("C Protein Distance Matrix")
}
```

```{r}
MatrixFunction(CproDM)
MatrixFunction(prMproDM)
MatrixFunction(eproDM)
```
## Group Vector Information and create Phylogeny

### Create a function to make a phylogenetic tree for proteins 
```{r}
# this functions takes in the protein for the tree to be created and whether it should create a phylogenetic tree for country or vector 

treefunction <- function(tree, type) {
  PhyloTree<- nj(tree)
  
  country = stri_extract_all(PhyloTree$tip.label, regex= "USA|Australia|South Africa|Israel|Russia|India|Madagascar|Azerbaijan|Ukraine|Czech Republic|Kenya|Nigeria|Central African Republic|DRC|Portugal|Morocco|Senegal|Tunisia|Cyprus|Japan|France", mode = "all", simplify =TRUE)  

  vect = stri_extract_all(PhyloTree$tip.label, regex= "Culex|Other|Equus|Homo sapiens|Corvus|Culiseta|Ornithodoros|Mus|Hyalomma|Oriolus|Aves|Dermacentor|Aedes|Rousettus|Culicidae",mode = "all", simplify =TRUE) 

if (type == "country"){
Groups = split(PhyloTree$tip.label, country) #Separate country from strain name
}
else if (type == "vect"){
Groups = split(PhyloTree$tip.label, vect) #Separate vector from strain name
}
col = groupOTU(PhyloTree,Groups) #Creates a group
ggtree(col, layout="rectangular", aes(color = group))#creates tree  
}

```

Using the phylogeny function we can create phylogenetic trees for the protein and method we select 
```{r}
#c protein trees
treefunction(CproDM, "country")
treefunction(CproDM, "vect")

#prM protein trees
treefunction(prMproDM, "country")
treefunction(prMproDM, "vect")

#e protein trees 
treefunction(eproDM, "country")
treefunction(eproDM, "vect")
```

```{r}
Cworld = read.csv("./input/Ccoord.csv")
rownames(Cworld) = Cworld[,1]
Cworld = subset(Cworld, select = -name)

Tree = nj(CproDM)
Tree$tip.label = stri_extract_all(Tree$tip.label, regex = "^.{8}",mode = "all", simplify =TRUE)
Tree = phytools::midpoint.root(Tree)

 phylo.to.map(Tree, Cworld,rotate = T)
``` 







