---
title: "data_visualization"
author: "Yifan Duan"
date: "01/04/2021"
output: html_document
---
BASH SCRIPT to generate cluster data
#!/bin/bash
sudo apt-get install cd-hit # install cd-hit; check github page for cdhit for installing on mac/windows
mkdir output/
cd-hit -i input/protein.fa -o cdhit_output/ -c 0.95 # run cdhit
cp cdhit_output/*.fas.1.clstr cluster.txt

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This code is used to visualize the data we will be working with to analyse West Nile Virus across different hosts and regions. 

```{r warning = FALSE}
#Loading required libraries 
library(ggplot2)
library(dplyr)
library(ggmap)
library(mapproj)
```

##Initializing the data 
```{r}
dat <- read.csv("./input/info.csv") #importing the data to use for visualizations. 

dat$length <- as.factor(dat$length) #making the length of the data a factor. 

dat <- mutate(dat, host = gsub("([A-Z].*) .*", "\\1", dat$host)) #using regex to isolate the host. 
dat <- mutate(dat, release_year = gsub("(\\d+)-.*", "\\1", dat$release_date)) #using regex to capture the release year.  
```

##Creating a World Map of WNV Frequency 
```{r}
country_df <- data.frame(table(dat$country)) #creates a data frame of all the countries. 
rownames(country_df) <- country_df$Var1 #re-names the country data frame row names in our data frame 
country_df$region <- rownames(country_df) #re-names the data frame regions by country 

world_map <- map_data("world") #
collection_map <- left_join(world_map, country_df, by = "region") #

# Create the map
svg(filename = './figures/location.svg') 
ggplot(collection_map, aes(long, lat, group = group))+
  geom_polygon(aes(fill = Freq), color = "white") +
  scale_fill_viridis_c(option = "C") #uses ggplot to create a world map of frequency 
dev.off()
```

##Release year visualization 
```{r}
# release year
svg(filename = './figures/release_year.svg')
ggplot(dat, aes(x = as.numeric(collection_date))) + geom_bar() +
  scale_x_continuous(name = "Collection date", breaks = seq(1950, 2000, by = 5)) #creates a histogram for the collection date of WNV 
dev.off()
```

## Genome length visualization 
```{r}
svg(filename = './figures/genome_length.svg')
ggplot(dat, aes(x = "", y = length, fill = length)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start = 0) + theme_void() + ggtitle("Genome Length") #creates a pie chart of the genome length ditribution 
dev.off()
```

## Vector Count visualization 
```{r}
host_count <- count(dat, host) #saves the count of host count in a new object 

common_name <- c("Unknown", "Mosquito (Aedes)", "Bird", "Crow", "Mosquito (Culex)", 
                 "Mosquito (Cuclicidae)", "Mosquito (Culiseta)", "Ticks (Dermacentor)", "Horse",
                 "Human", "Ticks (Hyalomma)", "Oriolus (Bird)", "Ticks (Ornithodoros)", "Fruit bat",
                 "Tissue culture")  #creates a vector of the vector common name 

species <- c("Unknown", "Mosquito", "Bird", "Bird", "Mosquito", "Mosquito", "Mosquito", "Ticks",
             "Horse", "Human", "Ticks", "Bird", "Ticks", "Bat", "Tissue") #creates a vector of all species names 

host_count <- cbind(host_count, common_name, Species = as.factor(species)) #combines the common names and species name vectors in a list 


svg(filename = './figures/host.svg')
host_count %>% ggplot(aes(x = reorder(common_name, n), y = n, fill = Species)) + 
  geom_col() + coord_flip() + 
  scale_x_discrete(name = "Species") + scale_y_continuous(name = "Count") #Creates a bar plot of the different species count 
dev.off()
```

##Protein percentage identity matrix
```{r}
data_CDS <- read.table('./input/pim.txt', header = F, sep = "") #reads data 
rnames <- data_CDS[,1] #indexes data to read only the first column 
data_CDS2 <- data.matrix(data_CDS[, 2:ncol(data_CDS)]) #creates a data matrix 
rownames(data_CDS2) <- rnames
colnames(data_CDS2) <- rnames

#Set the colour scheme
library(RColorBrewer) #loads the RColor Brewer package 
my_palette = colorRampPalette(c("white", "blue"))(n = 11) #creates colour scheme with desired colorus 

#Set other variables
library(gplots) #loads the gplots package 

#Prepare heatmaps
svg(filename = './figures/aai.svg')
my_palette = colorRampPalette(c("white", "dark blue"))(n = 5)
heatmap.2(data_CDS2, trace = c("none"), density.info = c("none"), 
          breaks = c(90, 92, 94, 96, 98, 100), col = my_palette, 
          lmat = matrix(c(4,2,3,1), nrow = 2, ncol = 2), key = T, labRow = rnames, labCol = rnames)
dev.off()
```

